\chapter{Preface}

This thesis presents original research conducted by the author under the supervision of Professor Karthik Pattabiraman in the Department of Electrical and Computer Engineering at the University of British Columbia.

\section*{Publications}

Portions of this thesis have been published or submitted for publication as follows:

\begin{enumerate}
    \item Yuqi Liu, Rui Xi, and Karthik Pattabiraman. ``Reentrancy Redux: The Evolution of Real-World Reentrancy Attacks on Blockchains.'' In \textit{Proceedings of the 55th IEEE/IFIP International Conference on Dependable Systems and Networks (DSN)}, 2025.

    \item Yuqi Liu, Rui Xi, and Karthik Pattabiraman. ``ReSect: A Tool for Automated Analysis of Reentrancy Transactions on Blockchains.'' Submitted to \textit{the 56th IEEE/IFIP International Conference on Dependable Systems and Networks (DSN)}, 2026.
\end{enumerate}

\section*{Author Contributions}

For the first paper (DSN 2025), I was responsible for data collection, qualitative case study, quantitative analysis, and manuscript writing. Rui Xi contributed to cross-validation of the analyses and manuscript revision. Professor Karthik Pattabiraman provided supervision, guidance on research direction, and manuscript revision.

For the second paper (DSN 2026), I was responsible for the design and implementation of \ToolName, dataset curation, experimental evaluation, and manuscript writing. Rui Xi contributed to manuscript revision. Professor Karthik Pattabiraman provided supervision and manuscript revision.

The thesis represents a synthesis and expansion of these works, presenting them as a unified research narrative with additional context, explanation, and analysis beyond what was possible in the conference paper format.

\section*{Ethics Approval}

This research did not require ethics approval as it involved only analysis of publicly available blockchain transaction data and did not involve human subjects.